---
title: "Disaster relief project - Part 1"
author:
  - Taeyoon Kim
  - Mauricio Mathey
  - Max Pearton
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---
```{r hide-code, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r}
#| message: FALSE
#| warning: FALSE

library(tidymodels)
library(tidyverse)
library(patchwork)
library(discrim)
library(kableExtra)
```

```{r}
#| message: FALSE
#| warning: FALSE

library(doParallel)
cl <- makePSOCKcluster(parallel::detectCores(logical = FALSE))
registerDoParallel(cl)

```



# Introduction

In response to the devastating earthquake that struck Haiti in 2010, rescue efforts
faced significant challenges in locating displaced persons residing in makeshift
shelters amidst the widespread destruction. With communication networks and
infrastructure severely disrupted, aid workers struggled to pinpoint the locations
of those in need amidst the vast expanse of affected areas. The presence of blue
tarps, used by displaced individuals to construct temporary shelters, served as
a vital indicator for locating those in need. The problem now wasn’t so much where
to look as it is how fast these unique locations could be pinpointed amidst a wide
array of geographical features and other man-made structures that are present in
high resolution geo-referenced imagery taken from an aircraft. The sheer quantity
of aerial footage collected on a daily basis deemed it an impossible task for
the human eye. Aid workers would be unable to search through the thousands of
images in time to correctly identify the tarps and subsequently coordinate rescue
operations. To address this critical issue, data mining techniques emerged as a
promising solution. Automated algorithms that are capable of efficiently and
accurately identifying these shelters would be deployed on the collected data,
enabling rescue workers to deliver essential aid in a timely manner. This project
aims to evaluate various classification algorithms to determine the most effective
approach for identifying displaced persons within the imagery data, with the
ultimate goal of facilitating timely assistance to those affected by the calamity.

Our endeavor began with extensive exploratory data analyses to gain a better
understanding of the underlying characteristics of the data collected by the
team from the Rochester Institute of Technology. The discoveries made in this
process informed our methodology in terms of how we manipulated the data, which
performance metric best served our interests, and how we justify recommending a
particular model given the task at hand.

# Data

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

data <- read.csv("~/MSDS/Statistical learning/Project/Part 1/HaitiPixels.csv")
```

We are working with data collected from aerial images. From Table 1, we can
see that the information that we have is class, red, green, blue. In our case,
as what we are trying to predict is if it's vegetation or a makeshift, Class
will be our dependent variable. Red, Green, and Blue are our independent
variables.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

kable(head(data), caption = 'Sample of the data')
```

From Table 2 we can see that there are 5 categories in the dataset. We are
interested in predicting the "Blue Tarp" category. A first issue that we can
observe is that it is an unbalanced set. Blue tarps represent 3% of the data only.
This means that it would be very easy for any method to ignore this class and it
could achieve a 97% accuracy. This suggests three initial strategies. The first
one is that when splitting the data we will need to take a stratified sample.
This will ensure that the distribution of both training and testing sets
resemble the original one. The second strategy is reducing the number of classes
from 5 to only 2. This will help us focus on identifying blue tarps and not
other types of objects. Finally, is the metric we will optimize for. If we choose
accuracy, it will be very easy for us to achieve an accuracy of 97%. But this
would be at the expense of not correctly identifying any shelters. On the other
hand if we are overly cautious and assign a greater amount of shelters, we could
potentially dilute rescue efforts. A metric that balances the sensitivity and
the specificity is the J-Index. This is the metric we will be using.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

table2 <- data %>% group_by(Class) %>% summarize(count = n()) %>%
  mutate(percentage = round(count/sum(count)*100,0))

kable(table2, caption = "Number of observations per class")
```
```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

data$Class2 <- ifelse(data$Class == 'Blue Tarp','Shelter','Other')
```

Let's explore how does the presence of red, blue, or green determine each class.
As previously mentioned, we will reclassify the data. Blue Tarp will be coded as
Shelter and all the other categories as Other. This will allow us to focus on
identifying only the shelters which is the problem we are trying to solve.

From figure 1 we can see that for red the shelters present most of the values
between ~120 and 225 which allow to differentiate them from other objects. For blue
we can see that shelters don't present low values of blue but they do present
values aboce 200 which could help differentiate. Finally for greens we observe
a similar behavior where shelter vs other exhibit differentiated behaviors.

```{r}
#| fig.cap: Distribution of color by class
#| fig.pos: H
#| fig.width: 5
#| fig.height: 5
#| fig.align: center
#| out.width: 75%

reds <- ggplot(data, aes(x = Red, color = Class2)) + geom_density() + theme_minimal()
blues <- ggplot(data, aes(color = Class2, x = Blue)) + geom_density() + theme_minimal()
greens <- ggplot(data, aes(color = Class2, x = Green)) + geom_density() + theme_minimal()

reds / blues / greens
```

Let's explore if there are any interactions between colors that could help us
identify shelters better.

From figure 2 we can see that there seems to be an approximately linear
separations between blue and red for shelter vs non shelter. This suggests that
methods with linear boundaries might perform well.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Blue vs Red separation
#| fig.pos: H
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| out.width: 75%
ggplot(data, aes(x = Red, y = Blue, color = Class2)) + geom_point() + theme_minimal()
```

From figure 3 we can see that while there is a separation between green and
red, it seems to be nonlinear. This because there is an overlaping area where
you would need a U shape function to be able to separe them. And even with that
we might not be able to do so.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Green vs Red separation
#| fig.pos: H
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| out.width: 75%

ggplot(data, aes(x = Red, y = Green, color = Class2)) + geom_point() + theme_minimal()
```

From figure 4 we can see that there seems to be a separation between green and
blue. While not exactly linear, it seems it could be approximated as one.


```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Green vs Blue separation
#| fig.pos: H
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| out.width: 75%

ggplot(data, aes(x = Blue, y = Green, color = Class2)) + geom_point() + theme_minimal()
```

# Description of methodology

We will test 5 different models: logistic regression, linear discriminant
analysis, quadratic discriminant analysis, K-nearest neighbor, and penalized
logistic regression. We will use 10 fold cross-validation to assess the
performance of each of the models and decide which one to use. A seed with a
value of 1 will be used across the models to ensure reproducibility. We will
have a training set and a testing set. We will get a stratified sample by the
class variable. The split will have an 80% - 20% split for training and
testing.



```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

set.seed(1)

# Set class as factor
data$Class2 <- as.factor(data$Class2)

# 2 Split and cross-validation
class_split <- initial_split(data, prop=0.80, strata=Class2)
train <- training(class_split)
test <- testing(class_split)

resamples <- vfold_cv(train, v=10, strata=Class2)
cv_control <- control_resamples(save_pred=TRUE)
```

## Logistic regression

The first model we are going to test is a logistic regression. From table 3
we can see that the performance of is quite good. We can achieve a sensitivity
of 99%, meaning that we are correctly identifying 99% of the shelters.
Additionally, we have a j-index of 0.89, indicating that we are correctly
discriminating non shelters from shelters.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Formula
formula <- Class2 ~ Red + Blue + Green

# Recipe
rec <- recipe(formula, data=train) %>% 
  step_normalize(all_numeric_predictors())

# Logistic regression specification
logreg_spec <- logistic_reg(engine="glm", mode="classification")

```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Logistic regression workflow
logreg_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(logreg_spec)
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Define metrics
metrics =  metric_set(roc_auc, j_index, sensitivity)

# Cross-validation
logreg_cv <- fit_resamples(logreg_wf, resamples,
                                metrics=metrics, control=cv_control)

```

```{r}
kable(collect_metrics(logreg_cv), caption = "Performance of logistic regression")
```

The next step is figuring out the optimal threshold. As previously mentioned,
we want to maximize j-index to prevent false positives.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Create function to scan thresholds
threshold_graph <- function(model_cv, model_name) {
  performance <- probably::threshold_perf(collect_predictions(model_cv),
                                          Class2, .pred_Shelter,
                                          thresholds=seq(0.05, 0.95, 0.01),
                                          event_level="second",
                                          metrics=metric_set(j_index))
  
  max_metrics <- performance %>%
    group_by(.metric) %>%
    filter(.estimate == max(.estimate))
  
  optimal_threshold <- max_metrics$.threshold[1]
  
  ggplot(performance, aes(x=.threshold, y=.estimate)) +
    geom_line() +
    geom_point(data=max_metrics, color="black") +
    labs(x="Threshold", y="J-Index") +
    geom_text(x = optimal_threshold, y = max_metrics$.estimate[1], 
              label = paste("Optimal Threshold: ", round(optimal_threshold, 2),
                            "|","J-Index: ", round(max_metrics$.estimate[1], 2)),
              vjust = 0.5, hjust = -0.3, color = "black", size = 4) +
    coord_cartesian(ylim=c(0, 1))
}
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Optimal threshold for logistic regression
#| fig.pos: H
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| out.width: 75%

threshold_graph(logreg_cv, "Logistic regression") + theme_minimal()
```

From figure 5 we can see that by reducing the threshold, the J-Index increases to 0.96.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

logreg_model <- fit(logreg_wf, data = train)
```

## Linear discriminant analysis

Next we are going to move to a linear discriminant analysis (LDA) model.

Before tuning the model we can see in table 4 a good performance. Just as with the logistic
regression. Nonetheless, the logistic regression performs slightly better. We
confirm this when we do the threshold scanning in figure 6 and identify that the highest
J-Index that LDA can achieve is 0.84 compared to 0.97. This suggest that
logistic regression is performing better.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Define LDA specifications
lda_spec <- discrim_linear(mode="classification") %>%
set_engine('MASS')

# LDA model specification
lda_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(lda_spec)

# Cross validation
lda_cv <- fit_resamples(lda_wf, resamples, metrics=metrics, control=cv_control)
```

```{r}
kable(collect_metrics(lda_cv), caption = "Performance of LDA")
```


```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Optimal threshold for LDA
#| fig.pos: H
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| out.width: 75%

threshold_graph(lda_cv, "LDA") + theme_minimal()
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

lda_model <- fit(lda_wf, data = train)
```


## Quadratic discriminant analysis

Next we are going to move to a quadratic discriminant analysis (QDA) model.

Before tuning the model we can see a good performance (table 5). QDA performs slightly
better than LDA but underperforms logistic regression. When tuning the threshold
in figure 7, the difference in performance vs LDA becomes very evident but it would still
underperform logistic regression.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Define LDA specifications
qda_spec <- discrim_quad(mode="classification") %>%
set_engine('MASS')

# LDA model specification
qda_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(qda_spec)

# Cross validation
qda_cv <- fit_resamples(qda_wf, resamples, metrics=metrics, control=cv_control)
```

```{r}
kable(collect_metrics(qda_cv), caption = "Performance of QDA")
```


```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Optimal threshold for QDA
#| fig.pos: H
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| out.width: 75%

threshold_graph(qda_cv, "QDA") + theme_minimal()
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

qda_model <- fit(qda_wf, data = train)
```

## K-nearest-neighbor

Next we will explore K-nearest-neighbors. For KNN we will define the number
of neighbors K as a tuneable parameter to understand which value would
yield the highest J-Index.

From table 6, we can see that if we aim to maximize the J-Index, the optimal number of
neighbors is 12. Other viable alternatives could be 9, 10, 11, or 13. All have
very similar J-Index values.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Define KNN specifications
knn_spec <- nearest_neighbor(engine='kknn',
                             mode="classification",
                             neighbors=tune())

# KNN model specification
knn_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(knn_spec)

# Define neighbors parameters
nn_params <- extract_parameter_set_dials(knn_wf) %>%
  update(neighbors=neighbors(c(2, 20)))


# Cross validation
knn_cv <- tune_grid(knn_wf,
                    metrics=metrics,
                    resamples=resamples,
                    control=cv_control,
                    grid=grid_regular(nn_params, levels=20))
```

```{r}
kable(show_best(knn_cv, metric='j_index'),
      caption = "Performance based on number of neighbors")
```

We can also see that with 12 neighbors we are not only maximizing the J-Index,
but that after this value AUC ROC doesn't seem to increase. In figure 8, we can
also observe that if we wanted to maximize sensitivity we would need 4 or 5 neighbors.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Optimal number of neighbors for KNN
#| fig.pos: H
#| fig.width: 5
#| fig.height: 5
#| fig.align: center
#| out.width: 75%

autoplot(knn_cv) + theme_minimal()
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

knn_model <- knn_wf %>%
  finalize_workflow(select_best(knn_cv, metric="j_index")) %>%
  fit(train)
```

## Penalized logistic regression

Finally, let's explore the results from penalized logistic regression. In this
case we have 2 tuning parameters. The first one is penalty, this determines
the amount of penalty that we will introduce in the model. The second one is
mixture. Which determines the proportion of Lasso vs Ridge penalization that
we will apply. 0 means only Ridge penalization is being used while 1 that only
Lasso. Anything in between is a mixture of both, with lower values having a
higher proportion of Ridge and higher ones a higher proportion of Lasso.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

# Set seed to ensure reproducibility
set.seed(1)

# Define penalized logistic regression specifications
penlog_spec <- logistic_reg(engine='glmnet',
                             mode="classification",
                             penalty=tune(),
                             mixture=tune())

# Penalized logistic regression model specification
penlog_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(penlog_spec)

# Define parameters
penlog_params <- extract_parameter_set_dials(penlog_wf) %>%
  update(penalty=penalty(c(-20, -5)),
         mixture=mixture(c(0, 1)))


# Cross validation
penlog_cv <- tune_grid(penlog_wf,
                    metrics=metrics,
                    resamples=resamples,
                    control=cv_control,
                    grid=grid_random(penlog_params, levels=500))
```

```{r}
kable(show_best(penlog_cv, metric='j_index'),
      caption = "Penalty and mixture tuning")
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Model performance based on regulazation and proportion of Lasso penalty
#| fig.pos: H
#| fig.width: 6
#| fig.height: 5
#| fig.align: center
#| out.width: 75%

autoplot(penlog_cv) + theme_minimal()
```

From both table 7 and the figure 9 we can see that the amount of penalty applied
is very small and that it is almost exclusively Lasso. It seems as if the
penalized logistic regression is performing worse than the other models.
Nonetheless, we see that there are high values for ROC AUC, which suggests that
0.5 might not be the optimal threshold.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE

penlog_model <- penlog_wf %>%
  finalize_workflow(select_best(penlog_cv, metric="j_index")) %>%
  fit(train)
```

In figure 10, we can see that just as with logistic regression, the optimal threshold is
0.05. which would yield a J-Index of 0.97.

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Optimal threshold for penalized logistic regression
#| fig.pos: H
#| fig.width: 6
#| fig.height: 5
#| fig.align: center
#| out.width: 75%

threshold_graph(penlog_cv, 'Penalized') + theme_minimal()
```

# Results

In figure 11 we can observe the ROC curve for logistic regression, penalized
logistic regression, LDA, and QDA. It is worth mentioning that KNN does not
have a traditional ROC as the other models. As we can see, the only models that
has a slightly worse ROC is LDA. The other three models are very similar.
To compare the performance of all the models we will look at accuracy, 
J-Index, sensitivity, and specificity.

```{r}
get_ROC_plot <- function(model, data, model_name) {
  model %>%
    augment(data) %>%
    roc_curve(truth=Class2, .pred_Shelter, event_level="second") %>%
    autoplot() + theme_minimal() +
    labs(title=model_name)
}
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: ROC for each model using testing data
#| fig.pos: H
#| fig.width: 6
#| fig.height: 5
#| fig.align: center
#| out.width: 75%

logreg_roc <- get_ROC_plot(logreg_model, test, "Logistic regression")
lda_roc <- get_ROC_plot(lda_model, test, "LDA")
qda_roc <- get_ROC_plot(qda_model, test, "QDA")
penlog_roc <- get_ROC_plot(penlog_model, test, "Penalized logistic regression")

(logreg_roc + penlog_roc) / (lda_roc + qda_roc)
```

From figure 12 we can see the performance metrics on testing data for each model.
As seen with the training data, all models have strong performance with KNN having
slightly better performance. We can observe that LDA's lower performance comes
from having a lower specificity. This means that LDA is more likely to identify
a shelter as a non-shelter, which would result is people not receiving help.

In our case, we are mostly interested in the J-Index and we can see that KNN
is slightly better than the others. As previously explained, J-Index is the
preferred metric because it allows us to balance both sensitivity and specificity
so rescue efforts are not diluted. From a practical standpoint nonetheless,
logistic regression is a close second and it has the advantage that
it is easier to interpret. Therefore our suggestion is to choose the logistic
regression model.

```{r}
calculate_metrics_probs <- function(model_name, model, data, threshold){
  results <- augment(model, data)
  results$prediction <- as.factor(ifelse(results$.pred_Shelter>threshold,'Shelter','Other'))
  metrics <- bind_rows(
   bind_cols(Model = model_name,j_index(results, Class2, prediction)),
   bind_cols(Model = model_name,accuracy(results, Class2, prediction)),
   bind_cols(Model = model_name,sensitivity(results, Class2, prediction)),
   bind_cols(Model = model_name,specificity(results, Class2, prediction)),
  )
  return(metrics)
}

calculate_metrics_knn <- function(model_name, model, data){
  results <- augment(model, data)
    metrics <- bind_rows(
      bind_cols(Model = model_name,j_index(results, Class2, .pred_class)),
      bind_cols(Model = model_name,accuracy(results, Class2, .pred_class)),
      bind_cols(Model = model_name,sensitivity(results, Class2, .pred_class)),
      bind_cols(Model = model_name,specificity(results, Class2, .pred_class)),
  )
  return(metrics)
}
```

```{r}
logreg_metrics <- calculate_metrics_probs('Logistic regression', logreg_model, test, 0.1)
lda_metrics <- calculate_metrics_probs('LDA', lda_model, test, 0.05)
qda_metrics <- calculate_metrics_probs('QDA', qda_model, test, 0.05)
penlog_metrics <- calculate_metrics_probs('Penalized', penlog_model, test, 0.05)
knn_metrics <- calculate_metrics_knn('KNN', knn_model, test)

total_metrics <- bind_rows(logreg_metrics, lda_metrics, qda_metrics,
                           penlog_metrics, knn_metrics)
```

```{r}
#| message: FALSE
#| warning: FALSE
#| cache: TRUE
#| fig.cap: Performance metrics for each model using testing data
#| fig.pos: H
#| fig.width: 6
#| fig.height: 5
#| fig.align: center
#| out.width: 75%

ggplot(total_metrics, aes(x = Model, y = .estimate)) + geom_col() +
  facet_wrap(~ .metric) + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_text(aes(label = round(.estimate,2)), vjust = 1.1,
            position = position_dodge(width = 0.9), color = "white")
```

# Conclusions

1. Determination of the Best Algorithm

After comprehensive evaluation of multiple classification algorithms, including
Logistic Regression, Linear Discriminant Analysis (LDA), Quadratic Discriminant
Analysis (QDA), K-Nearest Neighbors (KNN), and Penalized Logistic Regression, it
was found that all the models exhibited strong performance across various metrics.
KNN emerged as the top performer with a J-index of 0.97 closely followed by Logistic
Regression with a J-index of 0.96. Despite KNN’s slightly superior performance,
Logistic Regression offers the advantage of interpretability, making it a viable
alternative. Considering the balance between performance and interpretability,
our recommendation is to prioritize the logistic regression model for its strong
performance and ease of interpretation. 

2. Confidence in the Results and Practical Implications

Despite the nuanced differences in performance among the evaluated models, the
overall findings instill confidence in the efficacy of predictive modeling techniques
for locating displaced persons in disaster relief scenarios. By leveraging the
insights gained from this project and implementing the recommended strategies,
humanitarian organizations can enhance their ability to identify shelters with
displaced persons accurately, thereby facilitating timely and targeted assistance
to those in need. While uncertainties and limitations persist, the results
underscore the potential of predictive modeling to significantly impact and
improve disaster response efforts, potentially saving lives and mitigating the
impact of humanitarian crises.  

3. What additional recommended actions can be taken to improve results?

As is the case with many predictive modeling tasks on datasets, particularly in
the realm of geo-referenced imagery, the first and most evident action is to
establish a wider pipeline for collecting aerial footage. This action isn’t
necessarily taking issue with the volume of data available. Rather, this is more
of an effort to address the stark imbalance evident in the dataset with only 3%
of the data pertaining to blue tarp identifiers. The reasoning is that with a
larger quantity of data available to train predictive models, the better ability
of said model to generalize on unseen data. The second action is to verify that
the use of tarps specifically blue in color is the preferred identifier for those
in need of rescue. Although the scope of this project defines it as so, it is
difficult to presume that the abundance of blue tarp material should dictate
rescue efforts throughout the country. Perhaps exploring an alternate identifier
for survivors would be a more realistic assessment of the usefulness of data-mining
algorithms in an emergency setting.


# Appendix {-}
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```








